
# Why Automation?

```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.gd422c5de97_0_0")
```


All of science is built on results being reliable and continually working toward identifying more true/less wrong explanations about the world.

```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g27a5d68b667_0_1747")
```

The process of science is first repeatability -- can the same researcher with the same data get the same results? Undoubtedly in the early stages of an analysis, sometimes the results and output can be in flux. But as the analysis gets further polished and decisions are made, it should be that the same results can be obtained no matter how many times an analysis is run or re-run by the same researcher.

This brings us to the critical but previously historically overlooked part of the pyramid known as reproducibility. Reproducibility is what happens when another researcher can take the same data as the research #1 and get the same results. This is more difficult than it sounds at face value because data analysis requires so many decisions and variables. The order and ease of which something is re-run and the computing environment used to run the analysis are two such factors.

Keep in mind that consistent results (like those seen with reproducible work) are not automatically true but inconsistent results (like those seen with irreproducible work) cannot be true. In other words, **correctness is not the same as reproducibility but reproducibility is a necessary aspect of correctness.**

Reproducibility is the overlooked but critical step that allows replicability to happen. Replicability is when new data is collected that extends the findings of the first study. With this new data, hopefully the same type of analysis can be done that helps the field learn even more about the concepts that were learned in the first study.

### Why reproducibility is so important.

Reproducibility is not only important because all of science is built upon it but it also saves everyone time!

We can often underestimate the extent to which our work, code, and data are being used and reused by others in the scientific community. The extent to which our work is reproducible then, not only affect us and our immediate collaborates but could aid or hinder other researcher's work in an exponential scale.

```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g280d2b56f79_0_3916")
```

In other words, if 10 researchers reuse your work and all 10 of them spent 100 hours trying to get it to work **without** success, that's a lot of time to waste (10,000 hours)! But conversely, if your work was made with reproducibility aiding tools and skillsets (like automation that we are discussing in this course) then you could save other researchers loads of time! Let's say instead 9 out of 10 of the researchers who try to reproduce  your work (as opposed to running their own analysis from scratch) are able to do so in that time allotment, that saves them an insane amount of time and stress!


### Automation as a reproducibility tool

Automation is just one of many tools and skillsets that can aid the reproducibility of your work!


```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g27a5d68b667_0_1821")
```

Returning to those 10 researchers, if instead of having those 10 people manually try to reproduce our work every time we change it, what if we had robots do that work instead? That would not only help us re-run our results more quickly (because researchers are often busy) but also robots are much better at repetitive work. In other words, your human collaborator is great at many things but even your most reliable collaborator will not be as punctual as a robot who is programmed to do the job.


```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g27a5d68b667_0_1764")
```

### A real world example

Let's bring this into the terms of a very common story for science. Let's say you are a researcher who submitted a manuscript and a reviewer comes back and asks you to re run the analysis with a minor tweak; perhaps a parameter change.

```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g280d2b56f79_0_987")
```

If you developed your analysis without using reproducibility aiding practices and without automation, it is very likely that this seemingly simple task could take a lot of your time and brain power. Because while you don't think anything on your computer changed since you ran this analysis 6 months ago, your computing environment and the software it uses has been changing the entire time!

This kind of simple "this should be easy" situation can easily devolve into a huge rabbit hole -- when you thought this analysis was basically wrapped up.

```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g280d2b56f79_0_1003")
```

But, if you had been using the principles of CI/CD and reproducibility you may have a better chance that your analysis should still run reliably. And if it doesn't re-run reliably you will have more runs and set up that you'll be able to pull from to pinpoint where the bug is in your analysis re-run that is keeping it from running.

```{r, out.width = "100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g104d8c58b62_46_9")
```

By having automation keep tabs on your development, you will be less likely to be blind sided by bugs in a situation where you need to re-run your analysis (or adapt it for a new analysis!)
