
# GitHub Actions Fundamentals


```{r, fig.align='center', out.width="100%", echo = FALSE }
ottrpal::include_slide("https://docs.google.com/presentation/d/1YmwKdIy9BeQ3EShgZhvtb3MgR8P6iDX4DfFD65W_gdQ/edit#slide=id.gcc4fbee202_0_141")
```



Understand the basics of a GitHub Actions structure
Recognize computing environments as a key tool of reproducibility and automation
Navigate GitHub to find a details on a GitHub actions run


## GitHub Actions Basics


```{r, fig.align='center', out.width="100%", echo = FALSE }
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g280d2b56f79_0_2777")
```

All GitHub Actions involve answering three questions:

1. When should a thing run?
2. What should be run?
3. With what environment should the thing be run?

These questions and other specifications are set by writing a [YAML file](https://www.redhat.com/en/topics/automation/what-is-yaml). YAML files are human readable markup language files. Basically its a list that is easy for humans to read and write and computers can read them too. This makes it good for writing a GitHub Action. Essentially, we're going to write a YAML file to make a recipe that GitHub will read to know what/when/with what we are trying to do.

```{r, fig.align='center', out.width="100%", echo = FALSE }
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g280d2b56f79_0_2781")
```

The headlines about working with YAML files:

1. Everything is a list (kind of like a JSON file).
1. Indentions = subsets of a list
1. Spacing is VERY specific! -- incorrect spacing will definitely result in errors for your GitHub Action run.

Let's take a look at an example YAML.
Note that the what comes before a `:` is generally a name and indent indicate subsets of a list. So in the overall list of `food` we have sublists of `vegetables` and `fruits`. `#` can be used as a comment and will not be treated as code.

Additionally, `:` are often names. So `citrus` is the name for the item `oranges` and etc.
```
# A comment here which is ignored
food:
  - vegetables: tomatoes
  - fruits:
      citrus: oranges
      tropical: bananas
```


Two items that every GitHub Action YAML must contain is `on:` and `jobs:`.

- `on:` tells GitHub when something should be run. For example "whenever a pull request is opened".
- `jobs:` tells GitHub what should be run. For example "run this bash script".

```{r, fig.align='center', out.width="100%", echo = FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g280d2b56f79_0_2787")
```
