
# GitHub Action Variables

The GitHub Actions environments have variables that are already set by default in the environment but you can also set environment variables yourself.

### Default variables

There are variables that are set by default in the GitHub Actions environment can be handy to use in different instances.

To print things out, you can use this kind of notation:
One of many useful default variables we can grab is `GITHUB_REPOSITORY`. We can use it with bash underneath a `run:` like this:
```
echo ${GITHUB_REPOSITORY}
```

In this next exercise we'll explore different ways to use variables.

## Exercise 3

For this exercise, we are going to continue to use the example repository that we set up in the previous chapter.

1. Create a new branch to work from.

```{r, fig.align='center', out.width="100%", echo = FALSE }
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g2903858106e_0_4")
```
From command line:
```
`git checkout -b "env-var"`
```

2. For this exercise we are going to copy over another GHA yaml to explore. This time, move the `exploring-var-and-secrets.yml` file to your `.github/workflows` directories you made in the previous chapter.

```{r, fig.align='center', out.width="100%", echo = FALSE }
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g2903858106e_0_17")
```
From command line:
```
mv activity-1-sample-github-actions/exploring-var-and-secrets.yml .github/workflows/exploring-var-and-secrets.yml
```

3. Now follow the same set of steps we used in the previous chapter to Add, Commit, Push the changes.

```{r, fig.align='center', out.width="100%", echo = FALSE }
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g2903858106e_0_25")
```
From command line:
```
git add .github/*
git commit -m "exploring gha variables"
git push --set-upstream origin env-var
```

4. Now create a pull request with the changes you just made. (Refer to the previous chapter if you need reminders on how to do this).

5. On your pull request page on GitHub, click on the `Details` button next to your workflow run.

```{r, fig.align='center', out.width="100%", echo = FALSE }
ottrpal::include_slide("https://docs.google.com/presentation/d/1x0Cnk2Wcsg8HYkmXnXo_0PxmYCxAwzVrUQzb8DUDvTA/edit#slide=id.g280d2b56f79_0_3250")
```

### Set variables
